```{r include_packages_2, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis and also two functions
# used for labeling and referencing
if(!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")
if(!require(dplyr))
    install.packages("dplyr", repos = "http://cran.rstudio.com")
if(!require(ggplot2))
    install.packages("ggplot2", repos = "http://cran.rstudio.com")
if(!require(ggplot2))
    install.packages("bookdown", repos = "http://cran.rstudio.com")
if(!require(thesisdown)){
  library(devtools)
  devtools::install_github("ismayc/thesisdown")
  }
library(thesisdown)
```


# Results

We present here the results of all the implementations that were disclosed in the previous section. We perform numerical comparisons between all the methods, setting out differences accross number of simulations and timesteps. 

Heston [-@heston1993closed] gives a closed form used for comparison as the 'true' option value and enabling the results to be exposed in terms of bias ^[Defined as $\mathbb{E} \left[ \hat{\alpha} - \alpha \right]$] and RMSE (root mean square error) ^[Defined as $\sqrt{\mathbb{E}((\hat{\theta}-\theta)^2)}$]. 


The simulaton experiments were performed on a notebook with an Intel(R) Core(TM) i7-4500U CPU @ 1.80GHz processor and 8GB of RAM running on a linux x86_64 based OS, Fedora 25. Codes were all written in R an run on version 3.4.1 "Single Candle" [@rlang].

First of all, we chose a parametrization based on what we saw in other works, made some adjustments, like reducing the options' time to maturity due to the slower nature of R language and compared initial the results with the true option price given by the function *callHestoncf* belonging to the package NMOF [@nmofpack]. Parameters can be seen in Table below.




```{r param1, echo=FALSE, message=FALSE, results='asis'}
source("./r_code/parametrization.R")
```


\clearpage

To perform our simulations, we fixed a seed and saved the results in Table \@ref(res). Since the value given by the *callHestoncf* function with the parameters in Table \@ref(param) is 14.176, the method that best approached the "true" value was the modified (drift interpolated) exact algorithm with $100,000$ simulations. Altough the Euler scheme gives the same result (14.16) as the modified EA with $10,000$ simulations, it moves away from the closed form value when we run $100,000$ simulations. 


```{r results1, echo=FALSE, message=FALSE, results='asis'}
source("./r_code/results.R")
```

Results are observable in Figure \@ref(fig:modcomp1) also. The plot gives a good sense of possible biases associated with each method. To verify if in fact these methods present bias, we performed ten thousand simulations of ten thousand paths to the Euler, Kahl-Jackel (KJ) and drift interpolated exact algorithm (EA-DI) with 20 steps ($dt = 0.05$) . Clearly, from Figure \@ref(modcomp1), all three implementations produce bias (true option value is the black vertical line). Since this setup isn't ideal, we increased the steps to 100 ($dt = 0.01$) to see if our methods would converge. The methods histogram can be viewed in Figure \@ref(results10k001). 




```{r timmings1, echo=FALSE, message=FALSE, results='asis'}
source("./r_code/timmings.R")
```




```{r modcomp1, echo=FALSE, message=FALSE, results='asis', fig.asp = 0.7, fig.width = 5, fig.align='center', fig.cap='Comparison between models, 20 steps. \\label{modcomp1}'}
source("./r_code/mod_comp1.R")
```


```{r results10k, echo=FALSE, message=FALSE, results='asis', fig.asp = 1.2, fig.width = 5, fig.align='center', fig.cap='Comparison between models, 20 steps. \\label{results10k}'}
source("./r_code/results10k005.R")
```


```{r bias, echo=FALSE, message=FALSE, results='asis'}
source("./r_code/bias.R")
```


```{r modcomp2, echo=FALSE, message=FALSE, results='asis', fig.asp = 0.7, fig.width = 5, fig.align='center', fig.cap='Comparison between models, 20 steps. \\label{modcomp2}'}
source("./r_code/mod_comp2.R")
```

```{r results10k001, echo=FALSE, message=FALSE, results='asis', fig.asp = 1.2, fig.width = 5, fig.align='center', fig.cap='Comparison between models, 100 steps. \\label{results10k001}'}
source("./r_code/results10k001.R")
```