`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!--
If you feel it necessary to include an appendix, it goes here.
-->




# Black-Scholes formula {#bsformula}


In this appendix, we start from the following Geometric Brownian Motion process:


\begin{align*}
dS_t &= \mu S_t dt + \sigma S_t dB_t
\end{align*}


And we claim that the solution to this stochastic differential equation applying Itô's calculus is the following formula:

\begin{align*}
S_T &= S_0 \times \exp \left( \left( \mu  - \frac{\sigma^2}{2} \right) T  + \sigma B_T \right)
\end{align*}


\begin{proof}

If $S$ were deterministic, $dS_t/S_t$ would be the derivative of $\ln(S_t)$ with respect to $S$. This suggests to find an expression for the stochastic differential of $\ln(S_t)$, a function of the single random variable $S_t$.



\begin{align*} 
f(t,S) &= \ln(S) \\[10pt]
df(t,S) &= \frac{\partial f}{\partial S} dS + \frac{1}{2} \frac{\partial^2 f}{\partial S^2} (dS)^2 \\[10pt]
d\ln(S) &= \frac{d\ln(S)}{dS} dS + \frac{1}{2} \frac{d^2\ln(S)}{dS^2}(dS)^2 \\[10pt]
(dS)^2 &= \int_{0}^{t}{\left(\sigma \times S \right)^2} ds = \sigma^2 S^2 dt \\[10pt]
d\ln(S) &= \frac{1}{S} (\mu S dt + \sigma S dB) + \frac{1}{2}\frac{-1}{S^2} \sigma^2 S^2 dt \\[10pt]
d\ln(S) &= \left( \mu -  \frac{1}{2} \sigma^2 \right) dt + \sigma dB \\[10pt]
\int_{t=0}^{T}{d\ln(S_T)} &= \int_{t=0}^{T}{\left( \mu - \frac{\sigma^2}{2} \right) dt} + \int_{t=0}^{T}{\sigma dB_t} 
\end{align*}
\begin{align*} 
\ln(S_T) - \ln(S_0) &=  \left(\mu  - \frac{\sigma^2}{2} \right)  T + \sigma B_T \\[10pt]
\ln \left( \frac{S_T}{S_0} \right) &= \left( \mu  - \frac{\sigma^2}{2}  \right) T + \sigma B_T \\[10pt]
S_T &= S_0 \times \exp \left( \left( \mu  - \frac{\sigma^2}{2} \right) T  + \sigma B_T \right) 
\end{align*}
\end{proof}



# Bessel Function {#bessel}

The modified Bessel function of the first kind can be described in the shape of a contour integral (below) and is plotted for three different $\nu$ in Figure \@ref(fig:bessel).

\begin{align*} 
I_n(z)= \frac{1}{2 \pi i}\displaystyle \oint e^{(z/2)(t+1/t)}t^{-n-1}dt
\end{align*}

However, the Broadie-Kaya paper [-@broadie2006exact] presents it as a power series:

\begin{align*}
I_{\nu}(z)= \left(\frac{1}{2}z\right)^\nu \displaystyle \sum_{k=0}^{\infty}{\frac{\left(\frac{1}{4}z^2\right)^k}{\left(k!\Gamma \left(\nu+k+1\right)\right)}}
\end{align*}

With $\Gamma(x)$ is the gamma function and $\nu$ ia a complex number.



```{r bessel, echo=FALSE, message=FALSE, results='asis', fig.asp = 0.7, fig.width = 5, fig.align='center', fig.cap='Modified Bessel Functions of the First Kind \\label{bessel}'}
source("./r_code/bessel.R")
```  	



# Implementations

## Euler's {#eulerapp}

\small
```{r calle, eval=FALSE, echo=TRUE, size="scriptsize"}
hestoneuler <- function(S, X, r, v, theta, rho, k, 
                        sigma, t = 0, dt = NULL, tau = 1, N){

  if(is.null(dt)){ dt <- (tau-t)/1000}
  sequencia <- seq(t,tau,dt)
  ST <- matrix(NA, length(sequencia), N) #transformar em matrix
  aux <- NULL
  sqrt_dt <- sqrt(dt)

  for(i in sequencia){
    Zv <- stats::rnorm(N)
    Zt <- stats::rnorm(N)
    Zs <- rho * Zv + (sqrt(1 - (rho^2)) * Zt)
    aux <- v
    aux[v < 0] <- 0
    sqrt_aux <- sqrt(aux)
    S <- S * (1 + r * dt + sqrt_aux * Zs * sqrt_dt)
    S[S <= 0] = 0
    v <- v + k * dt * (theta - aux) + sigma * sqrt_aux * Zv * sqrt_dt
    ST[j,] <- S
  }
  rm(aux, v, Zv, Zt, Zs, S, j)
  ST <- as.matrix(ST, ncol=N)
  Result <- ST[nrow(ST),] - X
  Result[Result <= 0] = 0
  call = mean(exp(-r*(tau-t))*Result)
  lista = list('call' = call, 'Result' = Result, 'Spot' = ST)
  return(lista)
}
```


## Kahl-Jackel Implementation {#kjimp}

\footnotesize
```{r kj_imp, eval=FALSE, echo=TRUE, size="scriptsize"}
Hestoncallkj <- function(S, X, r, q, v, theta, rho, k, 
                         sigma, t = 0, dt = NULL, tau = 1, N){

    if(is.null(dt)){ dt <- (T-t)/1000}
    v <- rep(v,N)
    theta<- rep(theta,N)
    sequencia <- seq(t,tau,dt)
    ST <- matrix(NA, length(sequencia), N) #transformar em matrix
    S <- log(S)

    for(i in seq(t,tau,dt)){
        Zv <- stats::rnorm(N)
        Zt <- stats::rnorm(N)
        Zs <- rho * Zv + sqrt(1 - rho^2) * Zt
        vt <- (v + k * theta * dt + sigma * sqrt(v) * Zv * sqrt(dt) +
                  (1/4) * sigma^2 * dt * ((Zv)^2 - 1))/(1 + k * dt)
        vt[vt <= 0] <- v[vt <= 0] + k * dt * (theta[vt <= 0] - 
                        v[vt <= 0],0) + sigma * sqrt(v[vt <= 0],0) * 
                        Zv[vt <= 0] * sqrt(dt)
        v <- vt
        v[v<=0] <- 0
        vt[vt<=0] <- 0
        S <- S + (r - (v+vt)/4) * dt + rho * sqrt(v) * Zv * sqrt(dt) +
             (1/2) * (sqrt(v) + sqrt(vt)) * (Zs + rho * Zv) * sqrt(dt) +
             ((rho * sigma * dt)/2) * ((Zv)^2 - 1)
        S[S <= 0] = 0
        ST[j,] <- S
    }
    ST <- as.matrix(ST, ncol=N)
    Result <- exp(ST[nrow(ST),]) - X
    Result[Result <= 0] = 0
    call = mean(exp(-r*tau)*Result)

    lista = list('call' = call, 'Result' = Result, 'Spot' = ST)
    return(lista)
}
```



## Exact Algorithm {#eabk}

\footnotesize
```{r ea_bk, eval=FALSE, echo=TRUE, size="scriptsize"}
phi_heston <- function(a, v0, v_t, d){

    gamma_a <- sqrt(k^2 - 2 * sigma^2 * 1i*a)
    gammadt <- gamma_a * (tau-t)
    sqrtv0vt <- sqrt(v0*v_t)
    delta <- -k * (tau-t)

    part1 <- (gamma_a * exp(-(gamma_a - k)/2 * (tau-t)) * (1 - exp(delta)))/
        (k * (1- exp(- gammadt)))

    part2 <- exp((v0+v_t)/(sigma^2) * ( (k * (1 + exp(delta)))/(1-exp(delta)) -
                        (gamma_a * (1 + exp(- gammadt)))/(1-exp(- gammadt))))

    part3 <- Bessel::BesselI(z = ((4 * gamma_a * sqrtv0vt)/(sigma^2) *
                                      exp(- gammadt/2)/
                                      (1 - exp(- gammadt))), nu = 0.5*d - 1) /
        Bessel::BesselI(z = ((4 * k * sqrtv0vt)/(sigma^2) * (exp(delta/2))/
                                 (1-exp(delta))), nu = 0.5*d - 1)

    return (part1 * part2 * part3)
}



intv <- function(n, cf, v_t){
    integrand <- function(x, phi = cf){

        f2 <- function(u){
            Im(phi(u) * exp(-1i * u * x)) /u
        }
        return(f2)
    }

    ## integrate to "cdf"
    F_x <- function (x) {
        y <- 0.5 - 1/pi * integrate(integrand(x),  lower= 0, upper= 1000,
                                    rel.tol = 0.001, stop.on.error = FALSE)$value
        return(y)
    }


    ## endsign

    endsign <- function(f, sign = 1) {
        b <- sign
        while (sign * f(b) < 0) b <- 10 * b
        return(b)
    }

    ## inversion

    low_bound = -Inf
    upp_bound = Inf
    invcdf <- function(u) {
        subcdf <- function(t) F_x(t) - u
        if (low_bound == -Inf)
            low_bound <- endsign(subcdf, -1)
        if (upp_bound == Inf)
            upp_bound <- endsign(subcdf)
        return(uniroot(subcdf, lower=low_bound, upper=upp_bound, 
                       tol = 0.001220703)$root)
    }
    U <- stats::runif(n)
    sapply(U, invcdf)
}

hestonea_mod <- function(S, X, r, v, theta, rho, k, sigma, t = 0, tau = 1){

    d1 <- (4 * k * theta)/(sigma)^2
    c0 <- (sigma^2 * (1 - exp(-k*tau)))/(4*k)
    dt <- (tau-t)
    ST <- NULL

    # sampling V

    lambda <- (4*k*exp(-k*dt)*v)/(sigma^2 * (1-exp(-k*dt)))
    vt <- c0 * stats::rchisq(n = 1, df = d1, ncp = lambda)

    # Sampling int{V}

    phi <- function(a, v0=v, v_t=vt, d=d1){phi_heston(a, v0=v, v_t=vt, d=d1)}
    
    int_v <- intv(1, cf = phi, v_t=vt)
    # OR if you perform the drift interpolation scheme:
    #  int_v <- dt * ((1/2) * v + (1/2) * vt)
    
    
    # Sampling int{v}dw
    int_vdw <- (1/sigma) * (vt - v - k * theta * dt + k  * int_v)

    # Sampling S
    if( int_v >= 0){
        m <- log(S) + (r * (tau - t) - (1/2) * int_v + rho * int_vdw)
        std <- sqrt((1 - rho^2)) * sqrt(int_v)
        S <- exp(m + std * rnorm(1))
        v <- vt
        ST <- S
    } else {
        v <- vt
        ST <- rbind(ST,NA)}

    Result <- ST - X
    Result[Result <= 0] = 0
    call = exp(-r*tau)*Result
    lista = list('call' = call, 'Result' = Result, 'Spot' = ST)
    return(lista)
}
```


## Results with 20 steps {#resultsapp}




```{r results10k, echo=FALSE, message=FALSE, results='asis', fig.asp = 1.2, fig.width = 5, fig.align='center', fig.cap='Comparison between models, 20 steps. \\label{results10k}'}
source("./r_code/results10k005.R")
```


```{r timmings1, echo=FALSE, message=FALSE, results='asis', eval=FALSE}
source("./r_code/timmings.R")
```


```{r modcomp2, echo=FALSE, eval = FALSE, message=FALSE, results='asis', fig.asp = 0.7, fig.width = 5, fig.align='center', fig.cap='Comparison between models, 100 steps. \\label{modcomp2}'}
source("./r_code/mod_comp2.R")
```


\normalsize


\clearpage

<!-- **In the main Rmd file** -->



<!-- **In Chapter \@ref(ref-labels):** -->

<!-- ```{r ref.label='include_packages_2', results='hide', echo = TRUE} -->

<!-- ``` -->

