`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!--
If you feel it necessary to include an appendix, it goes here.
-->




# Black-Scholes formula {#bsformula}


In this appendix, we start from the following Geometric Brownian Motion process:


\begin{align*}
dS_t &= \mu S_t dt + \sigma S_t dB_t
\end{align*}


And we claim that the solution to this stochastic differential equation applying Itô's calculus is the following formula:

\begin{align*}
S_T &= S_0 \times \exp \left( \left( \mu  - \frac{\sigma^2}{2} \right) T  + \sigma B_T \right)
\end{align*}


\begin{proof}

If $S$ were deterministic, $dS_t/S_t$ would be the derivative of $\ln(S_t)$ with respect to $S$. This suggests to find an expression for the stochastic differential of $\ln(S_t)$, a function of the single random variable $S_t$.



\begin{align*} 
f(t,S) &= \ln(S) \\[10pt]
df(t,S) &= \cancelto{0}{\frac{\partial f}{\partial t}dt}  + \frac{\partial f}{\partial S} dS + \cancelto{0}{\frac{1}{2} \frac{\partial^2 f}{\partial t^2} (dt)^2} + \frac{1}{2} \frac{\partial^2 f}{\partial S^2} (dS)^2  + \cancelto{0}{\frac{\partial^2 f}{\partial t \partial S} dt dS} \\[10pt]
d\ln(S) &= \frac{d\ln(S)}{dS} dS + \frac{1}{2} \frac{d^2\ln(S)}{dS^2}(dS)^2 \\[10pt]
(dS)^2 &= \int_{0}^{t}{\left(\sigma \times S \right)^2} ds = \sigma^2 S^2 dt \\[10pt]
d\ln(S) &= \frac{1}{S} (\mu S dt + \sigma S dB) + \frac{1}{2}\frac{-1}{S^2} \sigma^2 S^2 dt \\[10pt]
d\ln(S) &= \left( \mu -  \frac{1}{2} \sigma^2 \right) dt + \sigma dB \\[10pt]
\int_{t=0}^{T}{d\ln(S_T)} &= \int_{t=0}^{T}{\left( \mu - \frac{\sigma^2}{2} \right) dt} + \int_{t=0}^{T}{\sigma dB_t} 
\end{align*}
\begin{align*} 
\ln(S_T) - \ln(S_0) &=  \left(\mu  - \frac{\sigma^2}{2} \right)  T + \sigma B_T \\[10pt]
\ln \left( \frac{S_T}{S_0} \right) &= \left( \mu  - \frac{\sigma^2}{2}  \right) T + \sigma B_T \\[10pt]
S_T &= S_0 \times \exp \left( \left( \mu  - \frac{\sigma^2}{2} \right) T  + \sigma B_T \right) 
\end{align*}
\end{proof}



# Bessel Function {#bessel}

The modified Bessel function of the first kind can be described in the shape of a contour integral (below) and is plotted for three different $\nu$ in Figure \@ref(fig:bessel).

\begin{align*} 
I_n(z)= \frac{1}{2 \pi i}\displaystyle \oint e^{(z/2)(t+1/t)}t^{-n-1}dt
\end{align*}

However, the Broadie-Kaya paper [-@broadie2006exact] presents it as a power series:

\begin{align*}
I_{\nu}(z)= \left(\frac{1}{2}z\right)^\nu \displaystyle \sum_{k=0}^{\infty}{\frac{\left(\frac{1}{4}z^2\right)^k}{\left(k!\Gamma \left(\nu+k+1\right)\right)}}
\end{align*}

With $\Gamma(x)$ is the gamma function and $\nu$ ia a complex number.



```{r bessel, echo=FALSE, message=FALSE, results='asis', fig.asp = 0.7, fig.width = 5, fig.align='center', fig.cap='Modified Bessel Functions of the First Kind \\label{bessel}'}
source("./r_code/bessel.R")
```  	




**In the main Rmd file**

 ```{r ref.label='include_packages', results='hide', echo = TRUE}
```

**In Chapter \@ref(ref-labels):**

<!-- ```{r ref.label='include_packages_2', results='hide', echo = TRUE} -->

<!-- ``` -->

